---
title: "Cross Correlation Analysis"
author: "Ileana Fenwick"
date: '2024-02-26'
output: html_document
---

---
title: "MHW_CCF_Final_Analysis"
author: "Ileana Fenwick"
date: '2023-05-04'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Desktop/Chapter 1/CTA_Chapter1")
library(ecodata)
library(ggplot2)
library(reshape2)
library(lattice)
library(maptools)
library(rgdal)
library(raster)
library(rgeos)
library(data.table)
library(vegan)
library(vegclust)
library(ecotraj)
library(survdat)
library(lwgeom)
library(tidyverse)
library(dplyr)
```

### 1.1 Wrangle traj metrics 

```{r Traj Metrics}
traj_metrics <- read_csv("final_trans_sp_traj_metrics.csv")
traj_metrics<-traj_metrics %>% 
              dplyr::select(EPU, YEAR, CTA_LEN, CTA_ANG )
traj_metrics<-traj_metrics %>% filter(YEAR > 1981)
#filtered for after 1982 because that is when the MHW time series begins. CCA requires matching data sets with no gaps. 
traj_metrics

#divide up by EPU 
mab_mets<-traj_metrics %>% filter(EPU == "MAB")
mab_mets

gom_mets<-traj_metrics %>% filter(EPU == "GOM")
gom_mets


gb_mets<-traj_metrics %>% filter(EPU == "GB")
gb_mets
```

### 1.2 Wrangle MHW from ECODATA reporting

```{r General MHW}
mhw<-ecodata::heatwave
mhw<-as.data.frame(mhw)
mhw
```

#### a) GOM 
```{r GOM MHW }

##subsetting for time series between 1982 to 2019 
cum_gom<-subset(mhw,EPU == "GOM" &  Var == "cumulative intensity" & Time > 1980 & Time < 2020)
cum_gom

#fill in years, any missing value means MHW cum int of 0 
cum_gom <-complete(cum_gom, Time = 1981:2019)
cum_gom<- as.data.frame(cum_gom)
#remove the EPU, Var, units columns 
#Units = * Celsius 
  cum_gom<-cum_gom[-c(2,4,5)]
cum_gom[is.na(cum_gom)]<-0


cum_gom<-cum_gom %>% rename(Cum_int = Value) %>% rename(YEAR = Time)
colnames(cum_gom)

#adding raw MHW cum int to final file 
for_gom_full_ccf<-cum_gom %>% subset(YEAR >1982)
for_gom_full_ccf
  
new_gomdiff<-data.frame(diff(as.matrix(cum_gom), lag = 1, differences = 1, na.omit = TRUE))
new_gomdiff
 new_gomdiff <-new_gomdiff[-c(1)]
new_gomdiff<-new_gomdiff %>% rename(Diff_MHW = Cum_int)

gom_ccfmhw<-data.frame(gom_mets, new_gomdiff)
##data frame with mhw data diff and traj metrics 
gom_ccfmhw

#data frame with trajectory metrics, diff in MHW cum int and raw MHW cum int 
full_gom_ccf<-right_join(gom_ccfmhw, for_gom_full_ccf, by = "YEAR")
full_gom_ccf

gom_ccf<-full_gom_ccf
gom_ccf
```

#### b) MAB 

```{r MAB MHW}

##subsetting for time series between 1982 to 2019 
cum_mab<-subset(mhw,EPU == "MAB" &  Var == "cumulative intensity" & Time > 1980 & Time < 2020)

#fill in years, any missing value means MHW cum int of 0 
cum_mab <-complete(cum_mab, Time = 1981:2019)

#remove 2017 
cum_mab<-cum_mab[-28,]

cum_mab
#fill in years, any missing value means MHW cum int of 0 

cum_mab<- as.data.frame(cum_mab)
#remove the EPU, Var, units columns 
#Units = * Celsius 
  cum_mab<-cum_mab[-c(2,4,5)]
cum_mab[is.na(cum_mab)]<-0

cum_mab<-cum_mab %>% rename(Cum_int = Value) %>% rename(YEAR = Time)
colnames(cum_mab)

#adding raw MHW cum int to final file 
for_mab_full_ccf<-cum_mab %>% subset(YEAR >1982)
for_mab_full_ccf
  
new_mabdiff<-data.frame(diff(as.matrix(cum_mab), lag = 1, differences = 1, na.omit = TRUE))
new_mabdiff
 new_mabdiff <-new_mabdiff[-c(1)]
new_mabdiff<-new_mabdiff %>% rename(Diff_MHW = Cum_int)

mab_ccfmhw<-data.frame(mab_mets, new_mabdiff)

##data frame with mhw data diff and traj metrics 
mab_ccfmhw

#data frame with trajectory metrics, diff in MHW cum int and raw MHW cum int 
full_mab_ccf<-right_join(mab_ccfmhw, for_mab_full_ccf, by = "YEAR")
full_mab_ccf

mab_ccf<-full_mab_ccf
mab_ccf

```

#### c) GB 
```{r GB MHW }

##subsetting for time series between 1982 to 2019 
cum_gb<-subset(mhw,EPU == "GB" &  Var == "cumulative intensity" & Time > 1980 & Time < 2020)
cum_gb

#fill in years, any missing value means MHW cum int of 0 
cum_gb <-complete(cum_gb, Time = 1981:2019)
cum_gb<- as.data.frame(cum_gb)
#remove the EPU, Var, units columns 
#Units = * Celsius 
  cum_gb<-cum_gb[-c(2,4,5)]
cum_gb[is.na(cum_gb)]<-0


cum_gb<-cum_gb %>% rename(Cum_int = Value) %>% rename(YEAR = Time)
colnames(cum_gb)

#adding raw MHW cum int to final file 
for_gb_full_ccf<-cum_gb %>% subset(YEAR >1982)
for_gb_full_ccf
  
new_gbdiff<-data.frame(diff(as.matrix(cum_gb), lag = 1, differences = 1, na.omit = TRUE))
new_gbdiff
 new_gbdiff <-new_gbdiff[-c(1)]
new_gbdiff<-new_gbdiff %>% rename(Diff_MHW = Cum_int)

gb_ccfmhw<-data.frame(gb_mets, new_gbdiff)
##data frame with mhw data diff and traj metrics 
gb_ccfmhw

#data frame with trajectory metrics, diff in MHW cum int and raw MHW cum int 
full_gb_ccf<-right_join(gb_ccfmhw, for_gb_full_ccf, by = "YEAR")
full_gb_ccf

gb_ccf<-full_gb_ccf
gb_ccf
```

### 2 Cross Correlation 
The CCF command is

ccf(x-variable name, y-variable name)

H is the correlation between the x variable at a time before t and the y variable at time t 

Mhw = x  
Traj lengths and angles = y 

H = negative, suggests x leads y 
H = positive, x lags y 

e.g. H = -5 and corr -0.45 means that MHW cum int (or diff) leads to a significantly lower traj length 5 years later 
the neg years is where x (MHW) occur before y (change) in time 

A correlation > 0.40 in the natural system would suggest something ecologically significant and worth testing further. In our work we formally use Yoo et al. 2023's significance threshold of 2/sqrt(n-|k|) where n is the total number of observations and k is the time lag. No correlations exceed this threshold 

Only negative time lags are considered ecologically relevant - MHW events LEAD community change metric shifts, cannot be the other way around and if there is the relationship is spurious.  


You will see me test two different cross correlations here (all insignificant) the first is the *difference* in MHW metrics from year to year. This felt like an important distinction and apples to apples comparison with CTA metrics because the lengths and angles are a measure of dissimilarity from year to year so we compare them to the difference in environmental variables from year to year. No singificant findings, reported in the supplementary figures of our MS. The second was the raw MHW data comparison to trajectory metrics. Also no significant findings and as such was excluded from the final MS. 

#### a) GOM - no signif 
```{r GOM CCFs}

##diff in MHW corr with CTA metrics 
gom_ccf_len<-ccf(gom_ccf$Diff_MHW, gom_ccf$CTA_LEN, na.action = na.pass,lag.max = 7, 
    ylab = "Cross Correlation",main = "lengths", ylim = c(-0.5,0.5))
gom_len_val<-ccf(gom_ccf$Diff_MHW, gom_ccf$CTA_LEN, na.action = na.pass,lag.max = 7)
gom_len_val
gom_ccf_ang<-ccf(gom_ccf$Diff_MHW, gom_ccf$CTA_ANG, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation",main = "angles", ylim = c(-0.5,0.5))
gom_ang_val<-ccf(gom_ccf$Diff_MHW, gom_ccf$CTA_ANG, na.action = na.pass,lag.max = 7)
gom_ang_val

#just trying out with MHW metric not diff, no significance 
gom_ccf_len_mhw_nodiff<-ccf(gom_ccf$Cum_int, gom_ccf$CTA_LEN, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation")

gom_ccf_ang_mhw_nodiff<-ccf(gom_ccf$Cum_int, gom_ccf$CTA_ANG, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation")

```


#### b) MAB - no signif 
```{r MAB CCFs}

##diff in MHW corr with CTA metrics 
mab_ccf_len<-ccf(mab_ccf$Diff_MHW, mab_ccf$CTA_LEN, na.action = na.pass,lag.max = 7, 
    ylab = "Cross Correlation",main = "lengths", ylim = c(-0.5,0.5))
mab_len_vals<-ccf(mab_ccf$Diff_MHW, mab_ccf$CTA_LEN, na.action = na.pass,lag.max = 7)
mab_len_vals

mab_ccf_ang<-ccf(mab_ccf$Diff_MHW, mab_ccf$CTA_ANG, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation",main = "angles", ylim = c(-0.5,0.5))
mab_ang_vals<-mab_ccf_ang<-ccf(mab_ccf$Diff_MHW, mab_ccf$CTA_ANG, na.action = na.pass,lag.max = 7)
mab_ang_vals

#just trying out with MHW metric not diff, no significance 
mab_ccf_len_mhw_nodiff<-ccf(mab_ccf$Cum_int, mab_ccf$CTA_LEN, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation")

mab_ccf_ang_mhw_nodiff<-ccf(mab_ccf$Cum_int, mab_ccf$CTA_ANG, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation")

```

#### a) GB - no signif 
```{r GB CCFs}

##diff in MHW corr with CTA metrics 
gb_ccf_len<-ccf(gb_ccf$Diff_MHW, gb_ccf$CTA_LEN, na.action = na.pass,lag.max = 7, 
    ylab = "Cross Correlation", ylim = c(-0.5,0.5), main = "lengths")
ccf(gb_ccf$Diff_MHW, gb_ccf$CTA_LEN, na.action = na.pass,lag.max = 7)
gb_len_ccfvalues = ccf(gb_ccf$Diff_MHW, gb_ccf$CTA_LEN, na.action = na.pass, lag.max = 7)
gb_len_ccfvalues

gb_ccf_ang<-ccf(gb_ccf$Diff_MHW, gb_ccf$CTA_ANG, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation", main = "angles", ylim = c(-0.5,0.5))

gb_ang_values<-ccf(gb_ccf$Diff_MHW, gb_ccf$CTA_ANG, na.action = na.pass,lag.max = 7)
gb_ang_values

#just trying out with MHW metric not diff, no significance 
gb_ccf_len_mhw_notdiff<-ccf(gb_ccf$Cum_int, gb_ccf$CTA_LEN, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation")

gb_ccf_ang_mhw_notdiff<-ccf(gb_ccf$Cum_int, gb_ccf$CTA_ANG, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation")

```


#### Exporting Figs 

```{r MAB figure export}
pdf(file = "/Users/ileanafenwick/Desktop/Chapter 1/Ch1 Publication Development/mab_ccf_plot_len.pdf", 
    width = 8,
    height = 8)

mab_ccf_len<-ccf(mab_ccf$Diff_MHW, mab_ccf$CTA_LEN, na.action = na.pass,lag.max = 7, 
    ylab = "Cross Correlation",main = " mab lengths", ylim = c(-0.5,0.5))

dev.off

pdf(file = "/Users/ileanafenwick/Desktop/Chapter 1/Ch1 Publication Development/mab_ccf_plot_ang.pdf", 
    width = 8,
    height = 8)

mab_ccf_ang<-ccf(mab_ccf$Diff_MHW, mab_ccf$CTA_ANG, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation",main = "mab angles", ylim = c(-0.5,0.5))

dev.off
```

```{r GOM figure export}

pdf(file = "/Users/ileanafenwick/Desktop/Chapter 1/Ch1 Publication Development/gom_ccf_plot_len.pdf", 
    width = 8,
    height = 8)

gom_ccf_len<-ccf(gom_ccf$Diff_MHW, gom_ccf$CTA_LEN, na.action = na.pass,lag.max = 7, 
    ylab = "Cross Correlation",main = " gom lengths", ylim = c(-0.5,0.5))

dev.off

pdf(file = "/Users/ileanafenwick/Desktop/Chapter 1/Ch1 Publication Development/gom_ccf_plot_ang.pdf", 
    width = 8,
    height = 8)

gom_ccf_ang<-ccf(gom_ccf$Diff_MHW, gom_ccf$CTA_ANG, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation",main = "gom angles", ylim = c(-0.5,0.5))

dev.off
```

```{r GB figure export}

pdf(file = "/Users/ileanafenwick/Desktop/Chapter 1/Ch1 Publication Development/gb_ccf_plot_len.pdf", 
    width = 8,
    height = 8)

gb_ccf_len<-ccf(gb_ccf$Diff_MHW, gb_ccf$CTA_LEN, na.action = na.pass,lag.max = 7, 
    ylab = "Cross Correlation",main = " gb lengths", ylim = c(-0.5,0.5))

dev.off

pdf(file = "/Users/ileanafenwick/Desktop/Chapter 1/Ch1 Publication Development/gb_ccf_plot_ang.pdf", 
    width = 8,
    height = 8)

gb_ccf_ang<-ccf(gb_ccf$Diff_MHW, gb_ccf$CTA_ANG, na.action = na.pass,lag.max = 7,
    ylab = "Cross Correlation",main = "gb angles", ylim = c(-0.5,0.5))

dev.off
```
