---
title: "Community Trajectory Analysis - Lengths and Angles"
author: "Ileana Fenwick"
date: '2024-02-20'
output: html_document
---

CTA-I outlines the methods for running a successful community trajectory analysis with lengths and angles illustrating community change from observation to observation. Look at CTA-II for the visualization of ordinations individually with reference points and extraction and visualization of distances from the starting observation. 
 
## Set up for analysis 

Vignette can be found https://github.com/NOAA-EDAB/survdat/blob/master/vignettes/calc_strat_mean.Rmd
May need to use --> remotes::install_github("NOAA-EDAB/survdat",build_vignettes=TRUE, force= TRUE)


# 1. Setup 
```{r Setup} 
#####Running these to get the datasets needed for plotting
setwd("~/Desktop/Chapter 1/CTA_Chapter1")
load("~/Desktop/Chapter 1/CTA_Chapter1/data/NEFSC_BTS_2021_all_seasons.RData")

library(reshape2)
library(lattice)
library(maptools)
library(sp)
library(rgdal)
library(raster)
library(rgeos)
library(data.table)
library(vegan)
library(permute)
library(vegclust)
library(ecotraj)
library(Rcpp)
library(survdat)
library(lwgeom)
library(dplyr)
library(remotes)
library(survdat)
library(data.table)


# install.packages("remotes")
# remotes::install_github("NOAA-EDAB/survdat")

```

#2. Data wrangling 
```{r Data formatting & mapping}
survdat <- survey$survdat
#quick check to make sure survdat covers areas of interest 
summary(survdat$YEAR)
# View(survdat)

######Plot the EPUs
areaPolygon <- sf::st_read(dsn = system.file("extdata","EPU.shp",package="survdat"), quiet=T)
plot_shapefile(areaPolygon)


##########Chunk
stations<-as.data.table(unique(survdat, by = c("CRUISE6", "STRATUM", "TOW")))
# View(stations)

data.table::setkey(stations, YEAR , STRATUM)
stations[, ntows := length(STATION), by = key(stations)]
head(stations)

```

These species were chosen based on a threshold of being caught in 0.01% of all surveys. Unclassified species are removed. 
```{r Defined species to include in analysis}

######The species
survdat<- survdat %>%
  filter(COMNAME == "ROUGHTAIL STINGRAY" |
           COMNAME == "ATLANTIC CROAKER" | 
           COMNAME == "BUTTERFISH"| 
           COMNAME == "LONGFIN SQUID" | 
           COMNAME == "SPOTTED HAKE"|
           COMNAME == "SEA SCALLOP" | 
           COMNAME == "NORTHERN SHORTFIN SQUID" |
           COMNAME == "SMOOTH DOGFISH"|
           COMNAME == "BULLNOSE RAY"|
           COMNAME == "ROUND HERRING"|
           COMNAME == "SCUP"|
           COMNAME == "NORTHERN SEAROBIN"| 
           # COMNAME == "TRASH"|##unclassified! include? 
           COMNAME == "BLUNTNOSE STINGRAY"| 
           COMNAME == "SPOT"|
           COMNAME == "SPINY BUTTERFLY RAY"|
           COMNAME == "COWNOSE RAY"|
           COMNAME == "LOGGERHEAD SEATURTLE"|
           COMNAME == "SANDBAR SHARK"|
           COMNAME == "FOURSPOT FLOUNDER"|
           COMNAME == "HORSESHOE CRAB"|
           COMNAME == "LITTLE SKATE"|
           COMNAME == "ATLANTIC ANGEL SHARK"|
           COMNAME == "AMERICAN LOBSTER"|
           #COMNAME == "DOGFISH,UNC"| ##unclassified! include? 
           COMNAME == "SOUTHERN STINGRAY"|
           COMNAME == "SUMMER FLOUNDER"|
           COMNAME == "BASKING SHARK"|
           COMNAME == "CHUB MACKEREL"|
           COMNAME == "BLUEFISH"|
           COMNAME == "ROSETTE SKATE"|
           COMNAME == "CLEARNOSE SKATE"|
           COMNAME == "GOOSEFISH"|
           # COMNAME == "RAY AND SKATE UNCL"|##unclassified! include? 
           COMNAME == "SILVER HAKE"|
           COMNAME == "SPINY DOGFISH"|
           COMNAME == "YELLOWTAIL FLOUNDER"|
           COMNAME == "WINDOWPANE"|
           COMNAME == "DUSKY SHARK"|
           #COMNAME == "SEA BASS UNCL"|##unclassified! include? 
           COMNAME == "BLACK SEA BASS"|
           COMNAME == "RED HAKE"|
           COMNAME == "CHAIN DOGFISH"|
           COMNAME == "OFFSHORE HAKE"|
           COMNAME == "STRIPED ANCHOVY"|
           COMNAME == "BLACKBELLY ROSEFISH"|
           COMNAME == "JOHN DORY"|
           COMNAME == "BEARDFISH"|
           COMNAME == "OCEAN SUNFISH"|
           COMNAME == "SOUTHERN EAGLE RAY"|
           COMNAME == "ATLANTIC ROCK CRAB"|
           COMNAME == "ATLANTIC SHARPNOSE SHARK"|
           COMNAME == "WEAKFISH"|
           COMNAME == "WINTER FLOUNDER"|
           COMNAME == "COBIA"|
           COMNAME == "FAWN CUSK-EEL"|
           COMNAME == "COARSEHAND LADY CRAB"|
           COMNAME == "ACADIAN REDFISH"|
           COMNAME == "HADDOCK"|
           COMNAME == "POLLOCK"|
           COMNAME == "ATLANTIC COD"|
           COMNAME == "WHITE HAKE"|
           COMNAME == "ATLANTIC HERRING"|
           COMNAME == "AMERICAN PLAICE"|
           COMNAME == "THORNY SKATE"|
           COMNAME == "WITCH FLOUNDER"|
           COMNAME == "SHRIMP UNCL"|
           COMNAME == "NORTHERN SHRIMP"|
           COMNAME == "ALEWIFE"|
           COMNAME == "CUSK"|
           COMNAME == "LONGHORN SCULPIN"|
           COMNAME == "WINTER SKATE"|
           COMNAME == "SMOOTH SKATE"|
           COMNAME == "BARNDOOR SKATE"|
           COMNAME == "SEA RAVEN"|
           COMNAME == "ATLANTIC MACKEREL"|
           COMNAME == "ATLANTIC WOLFFISH"|
           COMNAME == "ATLANTIC HALIBUT"|
           COMNAME == "OCEAN POUT"|
           COMNAME == "STRIPED BASS") 


length(unique(survdat$COMNAME)) ##76 species included in analysis

```


#3. Calculate stratified means


Stratified means for all EPUs 
```{r} 
area <- sf::st_read(dsn = system.file("extdata","EPU.shp",package="survdat"),quiet=T)
gbsurv<-calc_stratified_mean(survdat, 
                             areaPolygon = area, 
                             areaDescription = "EPU",
                             filterByArea = "GB", 
                             filterBySeason = "FALL", 
                             groupDescription = "COMNAME", 
                             filterByGroup = "all", 
                             mergesexFlag = T, 
                             tidy = F, 
                             returnPrepData = F
)

#add column to data set with the EPU in it 
gbsurv$EPU<-"GB"
###see how many unique years of data there are
length(unique(gbsurv$YEAR))#57 years

unique(gbsurv$YEAR)[order(unique(gbsurv$YEAR))]



gomsurv<-calc_stratified_mean(survdat, 
                              areaPolygon = area, 
                              areaDescription = "EPU",
                              filterByArea = "GOM", 
                              filterBySeason = "FALL", 
                              groupDescription = "COMNAME", 
                              filterByGroup = "all", 
                              mergesexFlag = T, 
                              tidy = F, 
                              returnPrepData = F
)

gomsurv$EPU<-"GOM"
length(unique(gomsurv$YEAR))#57 years
unique(gomsurv$YEAR)[order(unique(gomsurv$YEAR))]


mabsurv<-calc_stratified_mean(survdat, 
                              areaPolygon = area, 
                              areaDescription = "EPU",
                              filterByArea = "MAB", 
                              filterBySeason = "FALL", 
                              groupDescription = "COMNAME", 
                              filterByGroup = "all", 
                              mergesexFlag = T, 
                              tidy = F, 
                              returnPrepData = F
)
mabsurv$EPU<-"MAB"

#######NOTE: 2017 is missing in the MAB due to documented ship issues 
length(unique(mabsurv$YEAR))
unique(mabsurv$YEAR)[order(unique(mabsurv$YEAR))]


```

#4. Create full data frame to run analysis 

```{r} 
######Put it all together
sp_counts<-rbind(gbsurv, gomsurv, mabsurv)
fin<-dcast(sp_counts, EPU + YEAR ~ COMNAME, value.var = "strat.biomass")
fin<-fin %>% dplyr::filter(YEAR > 1968)
#could modify here is you need additional columns included in your final analysis data frame (e.g. SVSPP)
fin$EPU
fin$YEAR


fin[is.na(fin)]<-0#put 0s in where no catch occured in a given year
sp_data<-fin[,3:ncol(fin)]#just the species dataset no delimiter of EPU or other identifiers 
# View(sp_data)
```

#5. Ordination visualization 

For all three EPUs visualized on the same ordination figure (no connecting lines)

Visual of points in time using color as a  gradient for passage of time
```{r}
#Define some colors based on YEAR
cfunk<-colorRampPalette(c("yellow","red"))
fin$color<-cfunk(20)[as.numeric(cut(fin$YEAR,breaks=20))]
fin$mypch<-21

#or if you want to change the shape of the dot you can keep this code 
fin[fin$EPU == "GOM", "mypch"]<-22
fin[fin$EPU == "MAB", "mypch"]<-23
fin[fin$EPU == "GB", "mypch"]<-24

###############Create Distance matrix############
#note Hellinger transformation of community data (recommended for composition data)
H_E<-dist(decostand(sp_data, method = "hellinger",
                    na.rm = TRUE))
# View(H_E_ord)
H_E_ord<-cmdscale(H_E, add= TRUE, k = nrow(sp_data)-1)

#just black and white no color no shapes 
plot(H_E_ord$points[,1],H_E_ord$points[,2])

#color and shapes ordination 
plot(H_E_ord$points[,1],H_E_ord$points[,2],bg = fin$color, pch = fin$mypch, type = "p", col = "black",
     xlab = "Ordination Axis 1",
     ylab= "Ordination Axis 2",
     cex = 1) 


#Lets bind ordination axes to fin to make sure thigns stay aligned correctly
#colors but no shapes designated 
fin$ORD1<-H_E_ord$points[,1]#
fin$ORD2<-H_E_ord$points[,2]
plot(fin$ORD1, fin$ORD2, bg = fin$color, pch = 21,
     main = "ALL - Original plot")

```

#6. Running and visualizing the CTA 


```{r} 
#ordination with connected lines - CTA, you can define colors as needed, normally just black 
trajectoryPCoA(H_E, sites = fin$EPU,fin$YEAR, lwd = 1.5, traj.colors = c("turquoise", "red", "purple"), survey.labels = FALSE)

#GB was first, GOM second and MAB third in the order c(X,X,X)

#color blind friendly with smaller arrowheads 
trajectoryPCoA(H_E, sites = fin$EPU,fin$YEAR, lwd = 1.5, traj.colors = c("#E69F00", "#56B4E9", "#009E73"), length = 0.15, survey.labels = FALSE)
```

#7. Ordination with reference point comparison got to CTA-II

```{r}
#manuscript ready fig 


pdf(file = "/Users/ileanafenwick/Desktop/Chapter 1/Ch1 Publication Development/cta_plot.pdf", 
    width = 8,
    height = 8)

trajectoryPCoA(H_E, 
               sites = fin$EPU,fin$YEAR, 
               lwd = 1.5, 
               traj.colors = c("#56B4E9","#E69F00", "#009E73"),
               length = 0.15, 
               survey.labels = FALSE)

dev.off

```

