---
title: 'Redundancy Analysis'
author: "Ileana Fenwick"
date: '2024-02-20'
output: html_document
---
This markdown walks you through how to compute a redundancy analysis (RDA) with covariate observations and a species data frame per methods outlined in Fenwick et al. 2024. Code chunks here have been optimized so you will see it specifically walking through the Mid Atlantic Bight, and in our analysis this exact code was repeated for the Gulf of Maine (GOM) and Georges Bank (GB). Data for GB and the GOM are available in the repository. 

# General Prep and Loading 

Modify for your system. 
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(vegan)
library(readr)
library(data.table)

setwd("~/Desktop/Chapter 1/CTA_Chapter1")
```


```{r Read in and data frame setup for species data and env data, include = FALSE }

##NOTE sp data re already hellinger transformed
#read in pretransformed species data for each epu 
mab_sp <- read_csv("rda_data_clean/rda_mabsp.csv")
#38 x 76 means 1982 to 2019 
mab_sp<-as.data.frame(mab_sp)

##these variables were already standardized and centered for reading in the clean file 
mab_rda_variables<-read_csv("rda_data_clean/mab_rda_variables.csv")
mab_rda_variables

#removing bottom two rows because of missing var from zoops so analysis is 1982 to 2017 
#redundancy analysis requires a complete time series 
mab_rda_variables <-mab_rda_variables[-c(37,38),]
mab_sp<-mab_sp[-c(37,38),]
mab_sp_spring<-mab_sp_spring[-c(37,38),] 
#36 x 5 

#RENAMED FOR FIG PLOTTING 
# mab_rda_variables<-mab_rda_variables %>% rename('MHW CI' = MHW_cum_int)
# mab_rda_variables
```

# Testing global model significance 
### Fall 
```{r Global model testing } 

full_mab_fall<-rda(mab_sp~.,mab_rda_variables)
summary(full_mab_fall)
anova(full_mab_fall) #significant 
adjR2_full_fall<-RsquareAdj(full_mab_fall)$adj.r.squared
adjR2_full_fall # adj r2 that is explained by all 5 var is .0845

```

# Step-wise model selection 
Step-wise selection of environmental variables based two criteria: if their inclusion into the model leads to significant increase of explained variance, and if the AIC of the new model is lower than AIC of the more simple model. Does NOT consider as criteria whether the adjusted R2 of the model exceeds the adjusted R2 of the global model. 

### Fall 
Model selection includes MHW and Fogarty  
```{r Fall model selection }
mab_fall_0<-rda(mab_sp~1,data = mab_rda_variables) #model with only sp matrix and intercept
mab_fall_all<-rda(mab_sp~.,data = mab_rda_variables)
mabsel_fall<-ordistep(mab_fall_0, scope = formula (mab_fall_all), direction = 'forward')

mabsel_fall
#constrained = 0.015 

RsquareAdj(mabsel_fall)
#### adj r2 = .0774

#significance testing on RDA 
anova.cca(mabsel_fall, step = 1000)
summary(mabsel_fall)
####significant 

# fall_plot1<-ordiplot(sel_fall, type = "points", scaling = 1, cex = 1.5)
fall_plot2<-ordiplot(mabsel_fall, type = "points", scaling = 2, cex = 1.5)

```

## Also I explored this with other ordi2step code from https://r.qcbs.ca/workshop10/book-en/redundancy-analysis.html

```{r}
mabfwd.sel<-ordiR2step(mab_fall_0, 
                    scope = formula (mab_fall_all), 
                    direction = "forward", 
                    R2scope = TRUE, 
                    pstep = 1000, 
                    trace = TRUE)

mabstepwise_results <- summary(mabfwd.sel)

# Step: R2.adj= 0 
# Call: mab_sp ~ 1 
#  
#                    R2.adjusted
# <All variables>   0.0845069756
# + Fogarty         0.0436913266
# + MHW_cum_int     0.0436232241
# + mean_gsi        0.0253190660
# + bottom_temp_an  0.0102859327
# <none>            0.0000000000
# + zoop_ratio     -0.0000522687
# 
#           Df    AIC      F Pr(>F)   
# + Fogarty  1 -77.33 2.5991  0.002 **
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Step: R2.adj= 0.04369133 
# Call: mab_sp ~ Fogarty 
#  
#                  R2.adjusted
# <All variables>   0.08450698
# + MHW_cum_int     0.07740269
# + mean_gsi        0.06634261
# + bottom_temp_an  0.05516788
# + zoop_ratio      0.04450636
# <none>            0.04369133
# 
#               Df     AIC      F Pr(>F)   
# + MHW_cum_int  1 -77.696 2.2423  0.002 **
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Step: R2.adj= 0.07740269 
# Call: mab_sp ~ Fogarty + MHW_cum_int 
#  
#                  R2.adjusted
# <All variables>   0.08450698
# + bottom_temp_an  0.08056252
# + zoop_ratio      0.07924935
# <none>            0.07740269
# + mean_gsi        0.07508767
# 
#                  Df     AIC      F Pr(>F)
# + bottom_temp_an  1 -76.927 1.1134    0.3

mabfwd.sel$call
#rda(formula = mab_sp ~ Fogarty + MHW_cum_int, data = mab_rda_variables)


#saving this selected RDA for use moving forward 
mab_rda<-rda(mab_sp ~ Fogarty + MHW_cum_int, data = mab_rda_variables)
RsquareAdj(mab_rda)
# $r.squared
# [1] 0.1301225
# 
# $adj.r.squared
# [1] 0.07740269

## The explanatory variables Fogarty & MHW explain 7.7% of the variance in community compositon in the MAB 

anova.cca(mab_rda, step = 1000)
# Permutation test for rda under reduced model
# Permutation: free
# Number of permutations: 999
# 
# Model: rda(formula = mab_sp ~ Fogarty + MHW_cum_int, data = mab_rda_variables)
#          Df Variance      F Pr(>F)    
# Model     2 0.015047 2.4682  0.001 ***
# Residual 33 0.100588                  
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

anova.cca(mab_rda, step = 1000, by = "term") 
# Permutation test for rda under reduced model
# Terms added sequentially (first to last)
# Permutation: free
# Number of permutations: 999
# 
# Model: rda(formula = mab_sp ~ Fogarty + MHW_cum_int, data = mab_rda_variables)
#             Df Variance      F Pr(>F)    
# Fogarty      1 0.008212 2.6940  0.001 ***
# MHW_cum_int  1 0.006835 2.2423  0.001 ***
# Residual    33 0.100588                  
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

anova.cca(mab_rda, step = 1000, by = "axis") 
# Permutation test for rda under reduced model
# Forward tests for axes
# Permutation: free
# Number of permutations: 999
# 
# Model: rda(formula = mab_sp ~ Fogarty + MHW_cum_int, data = mab_rda_variables)
#          Df Variance      F Pr(>F)    
# RDA1      1 0.009272 3.0418  0.001 ***
# RDA2      1 0.005775 1.8946  0.008 ** 
# Residual 33 0.100588                  
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# >

#out full model is statistically significant (p = 0.001), every variable included in the model is significant, and every axis is signifcant. 

ordiplot(mab_rda, scaling = 2)

# Custom triplot code!

## extract % explained by the first 2 axes
perc <- round(100*(summary(mab_rda)$cont$importance[2, 1:2]), 2)

## extract scores - these are coordinates in the RDA space
sc_si <- scores(mab_rda, display="sites", choices=c(1,2), scaling=2)
sc_sp <- scores(mab_rda, display="species", choices=c(1,2), scaling=2)
sc_bp <- scores(mab_rda, display="bp", choices=c(1, 2), scaling=2)
years<-1982:2017
## Custom triplot, step by step

# Set up a blank plot with scaling, axes, and labels
plot(mab_rda,
     scaling = 2, # set scaling type 
     type = "none", # this excludes the plotting of any points from the results
     frame = FALSE,
     # set axis limits
     xlim = c(-1,1), 
     ylim = c(-1,1),
     # label the plot (title, and axes)
     main = "Triplot RDA - scaling 2",
     xlab = paste0("RDA1 (", perc[1], "%)"), 
     ylab = paste0("RDA2 (", perc[2], "%)") 
)
# add points for site scores -- these are the years in our rda 
points(sc_si,
       pch = 3, # set shape (here, circle with a fill colour)
       col = "red", # outline colour
       
       cex = 0.75) # size
# add points for species scores
text(sc_si + c(0.03, 0.09), # adjust text coordinates to avoid overlap with points 
     labels = years, 
     col = "grey40", 
     font = 2, # bold
     cex = 0.6)
points(sc_sp, 
       pch = 22, # set shape (here, square with a fill colour)
       col = "black",
       bg = "#f2bd33", 
       cex = 1.2)
# add text labels for species abbreviations
text(sc_sp + c(0.03, 0.09), # adjust text coordinates to avoid overlap with points 
     labels = rownames(sc_sp), 
     col = "grey40", 
     font = 2, # bold
     cex = 0.6)
# add arrows for effects of the expanatory variables
arrows(0,0, # start them from (0,0)
       sc_bp[,1], sc_bp[,2], # end them at the score value
       col = "red", 
       lwd = 3)
# add text labels for arrows
text(x = sc_bp[,1] -0.1, # adjust text coordinate to avoid overlap with arrow tip
     y = sc_bp[,2] - 0.03, 
     labels = rownames(sc_bp), 
     col = "red", 
     cex = 1, 
     font = 2)

# To create a color gradient based on years and apply it to the points in your triplot, you can use the `cfunk` color ramp palette that you defined. Here's how you can modify your code to include the color gradient:



```

FINAL PLOT FOR FIG IN MANUSCRIPT REVISION 
```{r} 
# Set up a blank plot with scaling, axes, and labels
pdf(file = "/Users/ileanafenwick/Desktop/revision_figs/mab_rda_revision.pdf.pdf", 
    width = 8,
    height = 8)


plot(mab_rda,
     scaling = 2,
     type = "none",
     frame = FALSE,
     xlim = c(-1, 1),
     ylim = c(-1, 1),
     main = "Mid Atlantic Bight",
     xlab = paste0("RDA1 (", perc[1], "%)"),
     ylab = paste0("RDA2 (", perc[2], "%)")
)

# Create a color gradient based on years using cfunk
color_gradient <- cfunk(length(years))

###Year scores color coded with legend red is later and yellow is earlier 
# Uncomment the following line to add points for site scores with color gradient
points(sc_si, pch = 21,
       col = "black", 
       bg = color_gradient, 
       cex = .85)

# add points for species scores
points(sc_sp,
       pch = 3, # set shape (here, square with a fill colour)
       col = "red",
       # bg = "#f2bd33",
       cex = 0.75)
# points(
#   jitter(sc_sp[, 1], factor = 0.25),  # Add jitter to x-axis
#   jitter(sc_sp[, 2], factor = 0.25),  # Add jitter to y-axis
#   pch = 3,  # set shape (here, square with a fill colour)
#   col = "red",
#   cex = 0.75
# )
arrows(0,0, # start them from (0,0)
       sc_bp[,1], sc_bp[,2], # end them at the score value
       col = "blue", 
       lwd = 1.5, 
       angle = 20, 
       length = 0.1)
# add text labels for arrows
text(x = sc_bp[,1] -0.1, # adjust text coordinate to avoid overlap with arrow tip
     y = sc_bp[,2] - 0.05, 
     labels = c("MHW CI", "Fogarty"), 
     col = "blue", 
     cex = 1, 
     font = 2)

box(col = "black", lwd = 2)


dev.off()  # Close the PDF device


```


# Variable partitioning 

Within the models that are selected with more than one variable, to account for a correlation between the two we can calculate the separation in variation of comm composition due to the correlated variables. 

More information : https://www.davidzeleny.net/anadat-r/doku.php/en:varpart_examples 


```{r variable partitioning for models with more than one selected variable}
# fractions [a+b+c]:
rda_all <- rda (mab_sp ~ Fogarty + MHW_cum_int, data = mab_rda_variables)
rda_all
anova(rda_all)
# fractions [a+b]:
rda_Fogarty <- rda (mab_sp ~ Fogarty, data = mab_rda_variables)
anova(rda_Fogarty)
# fractions [b+c]:
rda_mhw <- rda (mab_sp ~ MHW_cum_int, data = mab_rda_variables)
anova(rda_mhw)

#conditonal effect of fog
rda_fog_mhw<-rda(mab_sp ~ Fogarty + Condition (MHW_cum_int), data = mab_rda_variables)
anova(rda_fog_mhw)

#conditional effect mhw 
rda_mhw_fog<-rda(mab_sp ~ MHW_cum_int + Condition (Fogarty), data = mab_rda_variables)
anova(rda_mhw_fog)
```


```{r using varp for partitioning}
varp_mab_fall<-varpart(mab_sp, ~ Fogarty, ~ MHW_cum_int, data = mab_rda_variables) 

varp_mab_fall

plot(varp_mab_fall, digits = 2, Xnames = c('Fogarty', 'MHW Cum.Int.'), bg = c('yellow', 'blue')) 

```


## Final signif testing 
Now to test the significance of ALL RDAs and iterations 
```{r signif testing}

#global model - signif 
all_mab<-anova(rda_all)
library(report)
report(anova(rda_all))

anova(rda_all)

## Simple (marginal) effect of Fogarty (fraction [a+b]) - signif 
anova(rda_Fogarty)

## Simple (marginal) effect of mean_gsi (fraction [b+c]) - signif 
anova(rda_mhw)

## Conditional (partial) effect of dose (fraction [a]) - signif 
anova(rda_fog_mhw)

## Conditional (partial) effect of cover (fraction [c]) - signif 
anova(rda_mhw_fog)

```

From these results, you may see that all simple (marginal) and conditional (partial) effects of both predictors are significant at P < 0.05. 